Unsupervised embeddings capture latent structure of scientific mobility

### 2020-02-24 paper/current/main.tex
replaced proximity with distance

We also calculate two forms of proximity between organizations. 
The geographic distance, $g_{ij}$, is the pairwise geographic distance between organizations.
Geographic distance is calculated as the great circle distance, in kilometers, between pairs of organizations. 
When less than one kilometer, a distance of one is imputed. 
The embedding distance, $d_{ij}$, is calculated as the pairwise cosine distance between vectors from the learned organization embedding, where $v_{i}$ and $v_{j}$ represent the vectors and $d_{ij} = 1 - \frac{v_{i} \dot v_{j}}{\lVert v_{i} \rVert \lVert v_{j} \rVert}$.
\js{cosine similarity itself is not the distance}

@@

%
% Figure - Gravity law performance
%
\begin{figure}[h!]
	\centering
	\label{fig:gravity_performance}
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_performance.pdf}
	\caption{
		\textbf{Embedding distance outperforms and is more robust than geographic distance.}
		\textbf{a.} embedding distance (cosine distance between vectors) explains more of the flux between organization pairs than does geographic distance (\textbf{b}). 
		Red line is line of best fit. 
		Black dots are mean flux across binned proximity meas.
		Color indicates frequency of organization pairs within each hex bin.
		\textbf{c-d.} Predictions  of flux between organization pairs made using embedding distance out-perform those made using geographic distance. 
		Boxplots show distribution of actual flux for binned values of predicted flux.
		Box color corresponds to the degree to which the distribution overlaps $x = y$.
		\textbf{e-f.} Predictions made using model parameters learned from subset of same-country organization pairs. 
		Embedding distance outperforms those made for geographic distance.
		\textbf{g-h.} Embedding distance also outperforms geographic distance when predicting flux using model parameters estimated from only pairs of organizations in different countries. 
		Predictions are not meaningfully different if estimates are learned on all pairs of organizations (Fig. S\ref{fig:supp:gravity_prediction_subsets}).
	}
\end{figure}

@@

\subsection*{Embeddings capture latent geography and hierarchy}
%
% Figure - SemAxis results
%
\begin{figure}[hp!]
	\centering
	\label{fig:semaxis}
	\includegraphics[width=\textwidth]{\figdir/SemAxis/semaxis_fig.pdf}
	\caption{
		\textbf{Embedding captures latent geography and prestige hierarchy.}
		\textbf{a.} illustrates SemAxis projection along two axes, comparing California to Massachusetts universities (left to right), and between the top 20 and geographically-matched bottom 20 universities ranked by the 2018 Times Higher Education ranking of U.S. Universities (bottom to top). 
		Points correspond to organizations, and a sample of individual states are highlighted. 
		\textbf{b.} Comparison between the ranking of organizations by their rank in the Times ranking and the SemAxis rank derived from the similarity to the "Elite" end of the axis prestige axis \textit{a}.
		White points are those top and bottom 20 universities aggregated to define the ends of the axis. 
		The grey box corresponds to the top 20 and bottom 20 ranks. 
	  	The estimate of the spearman rank correlation between the SemAxis and Times ranking is 0.81 (n = 145).
	}
\end{figure}

%
% Subsection: Mathematical Exploration
%
\subsection*{Mathematical Exploration}

@@

@@

\newpage
%
% Figure - gravity correlation
%
\begin{figure}[p!]
	\centering
	\label{fig:supp:gravity_correlation}
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_correlations.pdf}
	\caption{
		\textbf{Embedding distance explains more variance for global, within, and between country flux than geographic distance.}
		\textbf{a.}
		Embedding distance (cosine distance between vectors) explains more of the total distance than does geographic distance (\textbf{b}). 
		Red line is line of best fit. 
		Black dots are mean flux across binned proximity meas.
		Color indicates frequency of organization pairs within each hex bin.
		Results here are identical to those shown in Fig.~\ref{fig:gravity_performance}a and Fig.~\ref{fig:gravity_performance}.
		\textbf{c-d.}	embedding distance explains more variance when considering only pairs of organizations within the same country. 
		\textbf{e-f.} embedding distance is more robust than geographic distance for pairs of organizations in different countries. 
	}
\end{figure}

@@

%
% Figure - PPR gravity model and predictions
%
\begin{figure}[p!]
	\centering
	\label{fig:supp:gravity_ppr}
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_ppr.pdf}
	\caption{
		\textbf{Personalized page rank does not capture mobility.}
		Performance of personalized page rank scores in explaining and predicting mobility.
		Personalized page rank is calculated for the underlying mobility network, and similarity between resulting vectors used as the PPR distance.
		PPR distance does not explain global mobility (\textbf{a}), or same (\textbf{b}) or different (\textbf{c}) country mobility, performing worse than geographic distance (Fig.~\ref{fig:supp:gravity_correlation}). 
		Red line is line of best fit. 
		Black dots are mean flux across binned proximity meas.
		Color indicates frequency of organization pairs within each hex bin.
		Similarly, PPR distance performs poorly as a distance metric for gravity model predictions globally (\textbf{d}), and across pairs of organizations in the same (\textit{e}) and in different (\textbf{f}) countries. 
		Boxplots show distribution of actual flux for binned values of predicted flux.
		Box color corresponds to the degree to which the distribution overlaps $x = y$.
	}
\end{figure}

>>>>

We also calculate two forms of distance between organizations. 
The geographic distance, $g_{ij}$, is the pairwise geographic distance between organizations.
Geographic distance is calculated as the great circle distance, in kilometers, between pairs of organizations. 
When less than one kilometer, a distance of one is imputed. 
The embedding distance, $d_{ij}$, is calculated as the pairwise cosine distance between vectors from the learned organization embedding, where $v_{i}$ and $v_{j}$ represent the vectors and $d_{ij} = 1 - \frac{v_{i} \dot v_{j}}{\lVert v_{i} \rVert \lVert v_{j} \rVert}$.

@@

%
% Figure - Gravity law performance
%
\begin{figure}[h!]
	\centering
	\label{fig:gravity_performance}
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_performance.pdf}
	\caption{
		\textbf{Embedding distance outperforms and is more robust than geographic distance.}
		\textbf{a.} embedding distance (cosine distance between vectors) explains more of the flux between organization pairs than does geographic distance (\textbf{b}). 
		Red line is line of best fit. 
		Black dots are mean flux across binned distances.
		Color indicates frequency of organization pairs within each hex bin.
		\textbf{c-d.} Predictions  of flux between organization pairs made using embedding distance out-perform those made using geographic distance. 
		Boxplots show distribution of actual flux for binned values of predicted flux.
		Box color corresponds to the degree to which the distribution overlaps $x = y$.
		\textbf{e-f.} Predictions made using model parameters learned from subset of same-country organization pairs. 
		Embedding distance outperforms those made for geographic distance.
		\textbf{g-h.} Embedding distance also outperforms geographic distance when predicting flux using model parameters estimated from only pairs of organizations in different countries. 
		Predictions are not meaningfully different if estimates are learned on all pairs of organizations (Fig. S\ref{fig:supp:gravity_prediction_subsets}).
	}
\end{figure}

@@

%
% Subsection: Mathematical Exploration
%
\subsection*{Mathematical explanation}

We observed that word embedding models trained on scientific mobility trajectories offers a reasonable measure of effective distance, one that explains flux between organizations far better than geographic distance. 
Why is this true?
Why does the skip-gram model learn vectors which approximate gravity law relationships?
We find evidence of a latent mathematical connection between the two models, one that makes the skip-gram model applicable to many other forms of mobility trajectory.

@@

%
% Subsection - hierarchy
% 
\subsection*{Embeddings capture latent geography and hierarchy}
%
% Figure - SemAxis results
%
\begin{figure}[hp!]
	\centering
	\label{fig:semaxis}
	\includegraphics[width=\textwidth]{\figdir/SemAxis/semaxis_fig.pdf}
	\caption{
		\textbf{Embedding captures latent geography and prestige hierarchy.}
		\textbf{a.} illustrates SemAxis projection along two axes, comparing California to Massachusetts universities (left to right), and between the top 20 and geographically-matched bottom 20 universities ranked by the 2018 Times Higher Education ranking of U.S. Universities (bottom to top). 
		Points correspond to organizations, and a sample of individual states are highlighted. 
		\textbf{b.} Comparison between the ranking of organizations by their rank in the Times ranking and the SemAxis rank derived from the similarity to the "Elite" end of the axis prestige axis \textit{a}.
		White points are those top and bottom 20 universities aggregated to define the ends of the axis. 
		The grey box corresponds to the top 20 and bottom 20 ranks. 
	  	The estimate of the spearman rank correlation between the SemAxis and Times ranking is 0.81 (n = 145).
	}
\end{figure}

%
% SECTION: DISCUSSION
% 
\section*{Discussion}

%
% SECTION: CONCLUSION
% 
\section*{Conclusion}

@@

\newpage
%
% Figure - gravity correlation
%
\begin{figure}[p!]
	\centering
	\label{fig:supp:gravity_correlation}
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_correlations.pdf}
	\caption{
		\textbf{Embedding distance explains more variance for global, within, and between country flux than geographic distance.}
		\textbf{a.}
		Embedding distance (cosine distance between vectors) explains more of the total distance than does geographic distance (\textbf{b}). 
		Red line is line of best fit. 
		Black dots are mean flux across binned distances.
		Color indicates frequency of organization pairs within each hex bin.
		Results here are identical to those shown in Fig.~\ref{fig:gravity_performance}a and Fig.~\ref{fig:gravity_performance}.
		\textbf{c-d.}	embedding distance explains more variance when considering only pairs of organizations within the same country. 
		\textbf{e-f.} embedding distance is more robust than geographic distance for pairs of organizations in different countries. 
	}
\end{figure}

@@

%
% Figure - PPR gravity model and predictions
%
\begin{figure}[p!]
	\centering
	\label{fig:supp:gravity_ppr}
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_ppr.pdf}
	\caption{
		\textbf{Personalized page rank does not capture mobility.}
		Performance of personalized page rank scores in explaining and predicting mobility.
		Personalized page rank is calculated for the underlying mobility network, and similarity between resulting vectors used as the PPR distance.
		PPR distance does not explain global mobility (\textbf{a}), or same (\textbf{b}) or different (\textbf{c}) country mobility, performing worse than geographic distance (Fig.~\ref{fig:supp:gravity_correlation}). 
		Red line is line of best fit. 
		Black dots are mean flux across binned distances.
		Color indicates frequency of organization pairs within each hex bin.
		Similarly, PPR distance performs poorly as a distance metric for gravity model predictions globally (\textbf{d}), and across pairs of organizations in the same (\textit{e}) and in different (\textbf{f}) countries. 
		Boxplots show distribution of actual flux for binned values of predicted flux.
		Box color corresponds to the degree to which the distribution overlaps $x = y$.
	}
\end{figure}
